---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# dosearch

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/santikka/dosearch/workflows/R-CMD-check/badge.svg)](https://github.com/santikka/dosearch/actions)
[![Codecov test coverage](https://codecov.io/gh/santikka/dosearch/branch/master/graph/badge.svg)](https://app.codecov.io/gh/santikka/dosearch?branch=master)

The `dosearch` [R](https://www.r-project.org/) package factilitates 
identification of causal effects from arbitrary observational and experimental 
probability distributions via do-calculus and standard probability 
manipulations using a search-based algorithm (Tikka et al., 2019, 2021).
Formulas of identifiable target distributions are returned in character format
using LaTeX syntax. The causal graph may additionally include mechanisms 
related to: 

  * Selection bias (Bareinboim and Tian, 2015)
  * Transportability (Bareinboim and Pearl, 2014) 
  * Missing data (Mohan et al., 2013)
  * Context-specific independence (Corander et al., 2019)

## Installation
You can install the latest release version from CRAN:
```{r, eval = FALSE}
install.packages("dosearch")
```

Alternatively, you can install the latest development version of `dosearch`:
```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("santikka/dosearch")
```

## Examples

```{r, echo = FALSE}
library(dosearch)
```

```{r, eval = TRUE}
# backdoor formula
data <- "p(x,y,z)"
query <- "p(y|do(x))"
graph <- "
  x -> y
  z -> x
  z -> y
"
dosearch(data, query, graph)

# frontdoor formula
graph <- "
  x -> z
  z -> y
  x <-> y
"
dosearch(data, query, graph)

# case-control design
data <- "
  p(x*,y*,r_x,r_y)
  p(y)
"
graph <- "
  x -> y
  y -> r_y
  r_y -> r_x
"
md <- "r_x : x, r_y : y"
dosearch(data, query, graph, missing_data = md)
```

## References

  * S. Tikka, A. Hyttinen, J. Karvanen, 
    **Causal effect identification from multiple incomplete data sources: a general search-based approach**, 
    *Journal of Statistical Software*, 
    99(5), 1--40, 2021.  ([published version](https://www.jstatsoft.org/index.php/jss/article/view/v099i05), [arXiv](https://arxiv.org/abs/1902.01073))
  * S. Tikka, A. Hyttinen, J. Karvanen, 
    **Identifying causal effects via context-specific independence relations**, 
    In *Proceedings of the 33rd Annual Conference on Neural Information Processing Systems*, 
    2019. ([published version](https://papers.nips.cc/paper/8547-identifying-causal-effects-via-context-specific-independence-relations), [arXiv](https://arxiv.org/abs/2009.09768))
  * E. Bareinboim and J. Tian,
    **Recovering causal effects from selection bias**,
    In *Proceedings of the Twenty-Ninth AAAI Conference on Artificial Intelligence*,
    2015. (http://ftp.cs.ucla.edu/pub/stat_ser/r445.pdf)
  * E. Bareinboim and J. Pearl,
    **Transportability from multiple environments with limited Experiments: completeness Results**,
    In *Advances of Neural Information Processing Systems 27*,
    2014. (http://ftp.cs.ucla.edu/pub/stat_ser/r443.pdf)
  * K. Mohan, J. Pearl, and J. Tian,
    **Graphical Models for Inference with Missing Data**,
    In *Advances of Neural Information Processing Systems 26*,
    2013. (http://ftp.cs.ucla.edu/pub/stat_ser/r410.pdf)
  * J. Corander, A. Hyttinen, J. Kontinen, J. Pensar, and J. Väänänen,
    **A logical approach to context-specific independence**,
    *Annals of Pure and Applied Logic*,
    170(9), 975--992, 2019. (https://doi.org/10.1016/j.apal.2019.04.004)
