---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# dosearch

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/santikka/dosearch/workflows/R-CMD-check/badge.svg)](https://github.com/santikka/dosearch/actions)
[![Codecov test coverage](https://codecov.io/gh/santikka/dosearch/branch/master/graph/badge.svg)](https://app.codecov.io/gh/santikka/dosearch?branch=master)
[![CRAN version](http://www.r-pkg.org/badges/version/dosearch)]( https://CRAN.R-project.org/package=dosearch)
<!-- badges::end -->

The `dosearch` [R](https://www.r-project.org/) package factilitates 
identification of causal effects from arbitrary observational and experimental 
probability distributions via do-calculus and standard probability 
manipulations using a search-based algorithm (Tikka et al., 2019, 2021).
Formulas of identifiable target distributions are returned in character format
using LaTeX syntax. The causal graph may additionally include mechanisms 
related to: 

* Selection bias (Bareinboim and Tian, 2015)
* Transportability (Bareinboim and Pearl, 2014) 
* Missing data (Mohan et al., 2013)
* Context-specific independence (Corander et al., 2019)
  
See the package vignette or the references for further information.
  
### Citing the package

If you use the `dosearch` package in a publication, please cite the corresponding
paper in the Journal of Statistical Software:

Tikka S, Hyttinen A, Karvanen J (2021). “Causal Effect Identification from 
Multiple Incomplete Data Sources: A General Search-Based Approach.” 
*Journal of Statistical Software*, 99(5), 1--40. 
[doi:10.18637/jss.v099.i05](https://doi.org/10.18637/jss.v099.i05).

## Installation

You can install the latest release version from CRAN:
```{r, eval = FALSE}
install.packages("dosearch")
```

Alternatively, you can install the latest development version of `dosearch`:
```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("santikka/dosearch")
```

## Examples

```{r, echo = FALSE}
library(dosearch)
```

```{r, eval = TRUE}
# backdoor formula
data <- "p(x,y,z)"
query <- "p(y|do(x))"
graph <- "
  x -> y
  z -> x
  z -> y
"
dosearch(data, query, graph)

# frontdoor formula
graph <- "
  x -> z
  z -> y
  x <-> y
"
dosearch(data, query, graph)

# case-control design
data <- "
  p(x*,y*,r_x,r_y)
  p(y)
"
graph <- "
  x -> y
  y -> r_y
  r_y -> r_x
"
md <- "r_x : x, r_y : y"
dosearch(data, query, graph, missing_data = md)
```

## References

* Tikka S, Hyttinen A, Karvanen J (2021).
  "Causal effect identification from multiple incomplete data sources: a general search-based approach." 
  *Journal of Statistical Software*, 
  99(5), 1--40.
  [doi:10.18637/jss.v099.i05](https://doi.org/10.18637/jss.v099.i05)
* Tikka S, Hyttinen A, Karvanen J (2019). 
  "Identifying causal effects via context-specific independence relations." 
  In *Proceedings of the 33rd Annual Conference on Neural Information Processing Systems*.
  (https://papers.nips.cc/paper/2019/hash/d88518acbcc3d08d1f18da62f9bb26ec-Abstract.html)
* Bareinboim E, Tian J (2015).
  "Recovering causal effects from selection bias." 
  In *Proceedings of the Twenty-Ninth AAAI Conference on Artificial Intelligence*,
  (http://ftp.cs.ucla.edu/pub/stat_ser/r445.pdf)
* Bareinboim E, Pearl J (2014).
  "Transportability from multiple environments with limited Experiments: completeness Results." 
  In *Advances of Neural Information Processing Systems 27*,
  (http://ftp.cs.ucla.edu/pub/stat_ser/r443.pdf)
* Mohan K, Pearl J, Tian J (2013).
  "Graphical models for inference with missing data." 
  In *Advances of Neural Information Processing Systems 26*,
  (http://ftp.cs.ucla.edu/pub/stat_ser/r410.pdf)
* Corander J, Hyttinen A, Kontinen J, Pensar J, Väänänen J (2019).
  "A logical approach to context-specific independence." 
  *Annals of Pure and Applied Logic*,
  170(9), 975--992. 
  [doi:10.1016/j.apal.2019.04.004](https://doi.org/10.1016/j.apal.2019.04.004)
